---
title: "RNA-seq Differential Expression and WGCNA Report"
subtitle: "DESeq2, enrichment, module preservation, and hub-gene summary"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
params:
  config_path: "../config/config.yaml"
---

```{r setup}
suppressPackageStartupMessages({
  library(yaml)
  library(knitr)
})

`%||%` <- function(a, b) if (is.null(a)) b else a

cfg <- yaml::read_yaml(params$config_path)
config_abs <- normalizePath(params$config_path, mustWork = TRUE)
repo_root <- dirname(dirname(config_abs))

is_abs_path <- function(path) {
  grepl("^[A-Za-z]:[\\\\/]", path) || startsWith(path, "/")
}

repo_path <- function(path, must_work = FALSE) {
  if (is.null(path) || is.na(path) || path == "") return(path)
  full_path <- if (is_abs_path(path)) path else file.path(repo_root, path)
  normalizePath(full_path, winslash = "/", mustWork = must_work)
}

path_or <- function(value, fallback) {
  if (is.null(value)) fallback else value
}

read_tsv <- function(path) {
  full_path <- repo_path(path)
  if (!file.exists(full_path)) return(data.frame())
  read.table(full_path, header = TRUE, sep = "\t", quote = "",
             stringsAsFactors = FALSE, check.names = FALSE)
}

fmt_p <- function(x, digits = 2) {
  x <- suppressWarnings(as.numeric(x))
  out <- ifelse(is.na(x), "",
                ifelse(x == 0, "<1e-300",
                       formatC(x, format = "e", digits = digits)))
  out
}

fmt_num <- function(x, digits = 2) {
  x <- suppressWarnings(as.numeric(x))
  ifelse(is.na(x), "", formatC(x, format = "f", digits = digits))
}

show_figures <- function(paths) {
  paths <- vapply(paths, repo_path, character(1))
  paths <- paths[file.exists(paths)]
  if (length(paths) == 0) {
    return(knitr::asis_output("_No figure file was found for this section._"))
  }
  knitr::include_graphics(paths)
}

show_table <- function(df, caption = NULL, digits = 3) {
  if (is.null(df) || nrow(df) == 0) {
    cat("_No table rows available._\n")
    return(invisible(NULL))
  }
  knitr::kable(df, caption = caption, digits = digits)
}

select_existing <- function(df, cols) {
  cols <- cols[cols %in% colnames(df)]
  df[, cols, drop = FALSE]
}

de_path <- path_or(cfg$paths$de_results_csv,
                   file.path(cfg$paths$tables_dir, "de_results.tsv"))
samples_path <- repo_path(cfg$paths$sample_metadata)
modules_path <- path_or(cfg$paths$module_assignments_csv,
                        file.path(cfg$paths$tables_dir,
                                  "module_assignments.tsv"))
go_path <- path_or(cfg$paths$enrichment_go_csv,
                   file.path(cfg$paths$tables_dir, "enrichment_go.tsv"))
kegg_path <- path_or(cfg$paths$enrichment_kegg_csv,
                     file.path(cfg$paths$tables_dir, "enrichment_kegg.tsv"))
gsea_go_path <- path_or(cfg$paths$gsea_go_csv,
                        file.path(cfg$paths$tables_dir, "gsea_go.tsv"))
gsea_kegg_path <- path_or(cfg$paths$gsea_kegg_csv,
                          file.path(cfg$paths$tables_dir, "gsea_kegg.tsv"))
module_ann_path <- path_or(cfg$paths$module_annotation_summary_csv,
                           file.path(cfg$paths$tables_dir,
                                     "module_annotation_summary.tsv"))
preservation_path <- path_or(cfg$paths$module_preservation_tsv,
                             file.path(cfg$paths$tables_dir,
                                       "module_preservation.tsv"))
hub_ranked_path <- path_or(cfg$paths$hub_genes_ranked_csv,
                           file.path(cfg$paths$tables_dir,
                                     "hub_genes_ranked.tsv"))
hub_priority_path <- path_or(cfg$paths$hub_genes_priority_csv,
                             file.path(cfg$paths$tables_dir,
                                       "hub_genes_priority_modules.tsv"))

de_path <- repo_path(de_path)
modules_path <- repo_path(modules_path)
go_path <- repo_path(go_path)
kegg_path <- repo_path(kegg_path)
gsea_go_path <- repo_path(gsea_go_path)
gsea_kegg_path <- repo_path(gsea_kegg_path)
module_ann_path <- repo_path(module_ann_path)
preservation_path <- repo_path(preservation_path)
hub_ranked_path <- repo_path(hub_ranked_path)
hub_priority_path <- repo_path(hub_priority_path)

de <- read_tsv(de_path)
samples <- read_tsv(samples_path)
modules <- read_tsv(modules_path)
go_tbl <- read_tsv(go_path)
kegg_tbl <- read_tsv(kegg_path)
gsea_go_tbl <- read_tsv(gsea_go_path)
gsea_kegg_tbl <- read_tsv(gsea_kegg_path)
module_ann <- read_tsv(module_ann_path)
preservation <- read_tsv(preservation_path)
hub_ranked <- read_tsv(hub_ranked_path)
hub_priority <- read_tsv(hub_priority_path)

fig_dir <- repo_path(cfg$paths$figures_dir)
```

# Executive Summary

This report summarizes the RNA-seq analysis pipeline for `r cfg$project$cancer %||% "the selected cohort"` (`r cfg$project$cancer_short %||% ""`). The analysis compares `r cfg$dataset$treatment_level %||% "treatment"` samples against `r cfg$dataset$reference_level %||% "reference"` samples, then uses WGCNA to find co-expression modules, annotate those modules, test preservation, and nominate hub-gene candidates.

In simple terms, the report follows this flow:

1. Differential expression asks which genes change between tumor and normal.
2. Enrichment asks what biological processes those changed genes represent.
3. WGCNA groups genes into modules that move together across samples.
4. Module annotation gives each module a biological label.
5. Preservation checks whether tumor modules are also present in normal tissue.
6. Hub-gene ranking finds central genes inside important modules.

```{r overview-table}
de_sig <- if (nrow(de) > 0) {
  !is.na(de$padj) &
    de$padj < cfg$deseq2$padj_threshold &
    abs(de$log2FoldChange) > cfg$deseq2$log2fc_threshold
} else {
  logical(0)
}

overview <- data.frame(
  Metric = c(
    "Samples",
    "Genes tested for DE",
    "Significant DE genes",
    "WGCNA genes assigned to modules",
    "Non-grey WGCNA modules",
    "Modules with annotation rows",
    "Modules with preservation calls",
    "Ranked hub rows",
    "Priority hub rows"
  ),
  Value = c(
    nrow(samples),
    nrow(de),
    sum(de_sig, na.rm = TRUE),
    nrow(modules),
    length(setdiff(unique(modules$module), "grey")),
    nrow(module_ann),
    nrow(preservation),
    nrow(hub_ranked),
    nrow(hub_priority)
  )
)
show_table(overview)
```

# Differential Expression

The DESeq2 step estimates how much each gene changes between tumor and normal. `log2FoldChange` is the size of the change; positive values mean higher in tumor, and negative values mean lower in tumor. `padj` is the multiple-testing corrected p-value.

```{r de-summary}
if (nrow(de) > 0) {
  up <- de_sig & de$log2FoldChange > cfg$deseq2$log2fc_threshold
  down <- de_sig & de$log2FoldChange < -cfg$deseq2$log2fc_threshold
  de_summary <- data.frame(
    Direction = c("Up-regulated", "Down-regulated"),
    Genes = c(sum(up, na.rm = TRUE), sum(down, na.rm = TRUE))
  )
  show_table(de_summary)
} else {
  cat("_DE results table was not found._\n")
}
```

```{r de-top-table}
if (nrow(de) > 0) {
  top_de <- de[de_sig, , drop = FALSE]
  top_de <- top_de[order(top_de$padj, -abs(top_de$log2FoldChange)),
                   , drop = FALSE]
  top_de <- head(top_de, 12)
  top_de$log2FoldChange <- fmt_num(top_de$log2FoldChange)
  top_de$padj <- fmt_p(top_de$padj)
  show_table(select_existing(top_de,
                             c("gene_symbol", "gene_id",
                               "log2FoldChange", "padj")),
             caption = "Top significant DE genes")
}
```

```{r de-figures, out.width="32%"}
show_figures(file.path(fig_dir,
                       c("ma_plot.png", "volcano.png",
                         "heatmap_top_de.png")))
```

# Functional Enrichment

Over-representation analysis tests the significant up- and down-regulated gene lists. GSEA uses the full ranked gene list, so it can detect biology spread across many genes even when individual genes do not pass a hard cutoff.

```{r enrichment-summary}
enrich_summary <- data.frame(
  Result = c("GO ORA rows", "KEGG ORA rows",
             "GO GSEA rows", "KEGG GSEA rows"),
  Rows = c(nrow(go_tbl), nrow(kegg_tbl),
           nrow(gsea_go_tbl), nrow(gsea_kegg_tbl))
)
show_table(enrich_summary)
```

```{r enrichment-figures, out.width="49%"}
show_figures(file.path(fig_dir,
                       c("go_bp_up.png", "go_bp_down.png",
                         "gsea_go_bp.png", "gsea_kegg.png")))
```

# WGCNA Co-expression Network

WGCNA groups genes into modules based on similar expression patterns. A module is a color label, not a biological label by itself. The biological meaning comes from module-trait correlations and enrichment results.

```{r module-size-table}
if (nrow(modules) > 0 && "module" %in% colnames(modules)) {
  module_sizes <- as.data.frame(sort(table(modules$module), decreasing = TRUE))
  colnames(module_sizes) <- c("Module", "Genes")
  show_table(head(module_sizes, 15), caption = "Largest WGCNA modules")
}
```

```{r wgcna-figures, out.width="49%"}
show_figures(file.path(fig_dir,
                       c("wgcna_sample_tree.png",
                         "wgcna_soft_threshold.png",
                         "wgcna_dendrogram.png",
                         "wgcna_module_trait.png")))
```

# Module Annotation

This section translates color modules into biological themes. The `best_padj` column is the strongest adjusted p-value among the top GO terms for that module.

```{r module-annotation-table}
if (nrow(module_ann) > 0) {
  module_ann_show <- module_ann
  module_ann_show$best_padj <- fmt_p(module_ann_show$best_padj)
  show_table(module_ann_show, caption = "Module annotation summary")
} else {
  cat("_Module annotation summary was not found._\n")
}
```

```{r module-enrichment-gallery, out.width="32%"}
module_figs <- file.path(fig_dir, "05_module_enrichment",
                         paste0("enrichment_",
                                c("turquoise", "brown", "blue",
                                  "cyan", "greenyellow", "black"),
                                ".png"))
show_figures(module_figs)
```

# Module Preservation

Module preservation compares the tumor network against normal samples. A high Z-summary means the module structure is preserved in normal tissue. A low Z-summary suggests tumor-specific network rewiring.

```{r preservation-table}
if (nrow(preservation) > 0) {
  preservation_show <- preservation
  if ("Zsummary" %in% colnames(preservation_show)) {
    preservation_show$Zsummary <- fmt_num(preservation_show$Zsummary)
  }
  show_table(select_existing(preservation_show,
                             c("module", "moduleSize",
                               "Zsummary", "medianRank", "class")),
             caption = "Module preservation calls")
}
```

```{r preservation-figures, out.width="49%"}
show_figures(file.path(fig_dir, "07_module_preservation",
                       c("preservation_zsummary.png",
                         "preservation_medianrank.png")))
```

# Hub-gene Candidates

Hub genes are genes with high module membership (`kME`) inside their own module. The priority table keeps modules that are trait-associated, not preserved, or strongly enriched.

```{r hub-summary}
if (nrow(hub_priority) > 0) {
  hub_summary <- aggregate(gene_id ~ module + priority_reason,
                           data = hub_priority, FUN = length)
  colnames(hub_summary) <- c("Module", "Priority reason", "Hub rows")
  hub_summary <- hub_summary[order(hub_summary$Module), ]
  show_table(hub_summary, caption = "Priority hub rows by module")
}
```

```{r hub-table}
if (nrow(hub_priority) > 0) {
  hub_show <- hub_priority[order(hub_priority$module,
                                 hub_priority$rank_in_module),
                           , drop = FALSE]
  hub_show <- head(hub_show, 30)
  hub_show$kME <- fmt_num(hub_show$kME)
  hub_show$log2FoldChange <- fmt_num(hub_show$log2FoldChange)
  hub_show$padj <- fmt_p(hub_show$padj)
  hub_show$enrichment_best_padj <- fmt_p(hub_show$enrichment_best_padj)
  show_table(select_existing(hub_show,
                             c("gene_symbol", "module", "kME",
                               "rank_in_module", "log2FoldChange",
                               "padj", "is_DE", "best_trait",
                               "trait_cor", "preservation_class",
                               "enrichment_best_padj",
                               "priority_reason")),
             caption = "First 30 priority hub-gene rows")
}
```

## Greenyellow Module Spotlight

The greenyellow module is preserved, but it is still biologically important because it has strong synaptic/neurotransmitter enrichment. In the hub-priority table this appears as `priority_reason = enriched`.

```{r greenyellow-table}
if (nrow(hub_priority) > 0) {
  gy <- hub_priority[hub_priority$module == "greenyellow", , drop = FALSE]
  if (nrow(gy) > 0) {
    gy <- gy[order(gy$rank_in_module), , drop = FALSE]
    gy$kME <- fmt_num(gy$kME)
    gy$log2FoldChange <- fmt_num(gy$log2FoldChange)
    gy$padj <- fmt_p(gy$padj)
    gy$enrichment_best_padj <- fmt_p(gy$enrichment_best_padj)
    show_table(select_existing(gy,
                               c("gene_symbol", "kME",
                                 "log2FoldChange", "padj",
                                 "is_DE", "preservation_class",
                                 "enrichment_best_padj",
                                 "priority_reason")),
               caption = "Greenyellow priority hub genes")
  } else {
    cat("_No greenyellow hub rows were present in the priority table._\n")
  }
}
```

```{r hub-figures, out.width="49%"}
show_figures(c(
  file.path(fig_dir, "hub_genes", "hubs_greenyellow.png"),
  file.path(fig_dir, "hub_genes", "gs_vs_mm",
            "gs_vs_mm_greenyellow_ER_status.png"),
  file.path(fig_dir, "hub_genes", "hubs_turquoise.png"),
  file.path(fig_dir, "hub_genes", "hubs_red.png")
))
```

# Output Files

The key machine-readable outputs are kept in `results/tables`, and the figures are kept in `results/figures`.

```{r output-files}
outputs <- data.frame(
  Output = c(
    "DE results",
    "GO enrichment",
    "KEGG enrichment",
    "GO GSEA",
    "KEGG GSEA",
    "Module assignments",
    "Module annotation summary",
    "Module preservation",
    "Ranked hubs",
    "Priority hubs"
  ),
  Path = c(
    de_path,
    go_path,
    kegg_path,
    gsea_go_path,
    gsea_kegg_path,
    modules_path,
    module_ann_path,
    preservation_path,
    hub_ranked_path,
    hub_priority_path
  )
)
outputs$Status <- ifelse(file.exists(outputs$Path), "found", "missing")
show_table(outputs)
```
